---
title: "Final Project"
author: 'Charlie White & William Nkemdirim'
date: "5/7/2021"
output: pdf_document
---

## Required Packages:

```{r, inclue=FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(SASxport))
suppressMessages(library(MASS))
suppressMessages(library(boot))
suppressMessages(library(ISLR))
suppressMessages(library(leaps))
suppressMessages(library(zoo))
suppressMessages(library(ROCR))
suppressMessages(library(randomForest))
suppressMessages(library(gbm))
suppressMessages(library(tree))
suppressMessages(library(e1071))
suppressMessages(library(glmnet))
suppressMessages(library(class))
suppressMessages(library(ROCR))
suppressMessages(library(PRROC))
```

## Data Sources

### HSQ_J, DBQ_J, DUQ_J, DPQ_J, SMQ_J Datasets

https://wwwn.cdc.gov/nchs/nhanes/Search/DataPage.aspx?Component=Questionnaire&CycleBeginYear=2017

### OHXDEN Oral Health Dataset

https://wwwn.cdc.gov/nchs/nhanes/Search/DataPage.aspx?Component=Examination&CycleBeginYear=2017

## Installing Datasets:

```{r}
setwd("/Users/charliewhite/Downloads")
DPQ_J <- read.xport("DPQ_J.XPT.txt")
OHXDEN <- read.xport("OHXDEN_J.XPT.txt")
SMQ_J <- read.xport("SMQ_J.XPT.txt")
DBQ_J <- read.xport("DBQ_J.XPT.txt")
HSQ_J <- read.xport("HSQ_J.XPT.txt")
DUQ_J <- read.xport("DUQ_J.XPT.txt")
```

## Data Cleaning and Pre-processing Steps

```{r}
which( colnames(DUQ_J)=="DUQ200")
which( colnames(DUQ_J)=="DUQ240")
which( colnames(DUQ_J)=="DUQ370")
# dropping unsused columns
DUQ_J <- DUQ_J[,c(1,2,13,31)]
DUQ_J[DUQ_J=="7"] <- NA
DUQ_J[DUQ_J=="9"] <- NA
DUQ_J$DUQ200 <- ifelse(DUQ_J$DUQ200 == 1, 1, 0)
DUQ_J$DUQ240 <- ifelse(DUQ_J$DUQ240 == 1, 1, 0)
DUQ_J$DUQ370 <- ifelse(DUQ_J$DUQ370 == 1, 1, 0)
DUQ_J <- DUQ_J %>% drop_na()
# creating a combined drug use variable
DUQ_J$DCOMBO <- DUQ_J$DUQ200 + DUQ_J$DUQ240 + DUQ_J$DUQ370
table(DUQ_J$DCOMBO)

# assigning and dropping missing values
which( colnames(HSQ_J)=="HSD010")
HSQ_J <- HSQ_J[,c(1,2)]
HSQ_J[HSQ_J=="7"] <- NA
HSQ_J[HSQ_J=="9"] <- NA
HSQ_J <- HSQ_J %>% drop_na()
table(HSQ_J$HSD010)

which( colnames(DBQ_J)=="DBQ700")
DBQ_J <- DBQ_J[,c(1,14)]
DBQ_J[DBQ_J=="7"] <- NA
DBQ_J[DBQ_J=="9"] <- NA
DBQ_J <- DBQ_J %>% drop_na()
table(DBQ_J$DBQ700)

SMQ_J <- SMQ_J[,c(1:2)]
SMQ_J$SMQ020 <- ifelse(SMQ_J$SMQ020 == 1, 1, 0)
SMQ_J <- SMQ_J %>% drop_na()
table(SMQ_J$SMQ020)
# a value of 0 indicates person has not smoked at least 100 cigarettes in their lifetime
# a value of 1 indicates they have smoked at least 100 cigarettes in their lifetime

which( colnames(OHXDEN)=="OHX31CTC" )
OHXDEN <- OHXDEN[,c(1:64)]
OHXDEN[] <- lapply(OHXDEN, as.character)
DPQ_J[] <- lapply(DPQ_J, as.character)
DPQ_J$SEQN <- as.character(DPQ_J$SEQN)

# assigning a 1 to teeth that required medical attention and a 0 to sound permanent/primary teeth
OHXDEN[OHXDEN=="A"] <- 1
OHXDEN[OHXDEN=="D"] <- 0
OHXDEN[OHXDEN=="E"] <- 1
OHXDEN[OHXDEN=="F"] <- 1
OHXDEN[OHXDEN=="J"] <- 1
OHXDEN[OHXDEN=="K"] <- 1
OHXDEN[OHXDEN=="M"] <- 1
OHXDEN[OHXDEN=="P"] <- 1
OHXDEN[OHXDEN=="Q"] <- 1
OHXDEN[OHXDEN=="R"] <- 1
OHXDEN[OHXDEN=="S"] <- 0
OHXDEN[OHXDEN=="T"] <- 1
OHXDEN[OHXDEN=="U"] <- 1
OHXDEN[OHXDEN=="X"] <- 1
OHXDEN[OHXDEN=="Y"] <- 1
OHXDEN[OHXDEN=="Z"] <- 1
DPQ_J[DPQ_J=="7"] <- NA
DPQ_J[DPQ_J=="9"] <- NA

which( colnames(OHXDEN)=="OHX01TC")
which( colnames(OHXDEN)=="OHX32TC")


# assigning a 1 to sound permanent/primary teeth and a 0 to implants/missing teeth
OHXDEN[ , 5:36][ OHXDEN[ , 5:36] == 1] <- 1
OHXDEN[ , 5:36][ OHXDEN[ , 5:36] == 2] <- 1
OHXDEN[ , 5:36][ OHXDEN[ , 5:36] == 3] <- 0
OHXDEN[ , 5:36][ OHXDEN[ , 5:36] == 4] <- 0
OHXDEN[ , 5:36][ OHXDEN[ , 5:36] == 5] <- 0

# combining all datasets

df <- left_join(DPQ_J, OHXDEN, by = "SEQN")
df <- df %>% drop_na()
which( colnames(df)=="OHDEXSTS")
which( colnames(df)=="OHDDESTS")
df <- df[, -c(12,13)]
df[] <- lapply(df, as.numeric)
SMQ_J[] <- lapply(SMQ_J, as.numeric)
df <- left_join(df, SMQ_J)
df <- df %>% drop_na()
DBQ_J[] <- lapply(DBQ_J, as.numeric)
df <- left_join(df, DBQ_J)
df <- df %>% drop_na()
HSQ_J[] <- lapply(HSQ_J, as.numeric)
df <- left_join(df, HSQ_J)
df <- df %>% drop_na()
DUQ_J[] <- lapply(DUQ_J, as.numeric)
df <- left_join(df, DUQ_J)
df <- df %>% drop_na()
df[] <- lapply(df, as.numeric)
```

## Generating Relevant Variables and Calculating Descriptive Statistics

```{r}
#creating DPQ_score from mental health questionnaire 
df$DPQ_score <- (df$DPQ010 + df$DPQ020 + df$DPQ030 + df$DPQ040 + df$DPQ050 + df$DPQ060 + df$DPQ070 + df$DPQ080 + df$DPQ090 + df$DPQ100)
table(df$DPQ_score)
summary(df$DPQ_score)
sd(df$DPQ_score)
IQR(df$DPQ_score)

#creating number of healthy permanent teeth variable
df$num_teeth <- (df$OHX01TC + df$OHX02TC + df$OHX03TC + df$OHX04TC + df$OHX05TC + df$OHX06TC + df$OHX07TC + df$OHX08TC +  df$OHX09TC + df$OHX10TC + df$OHX11TC + df$OHX12TC +  df$OHX13TC + df$OHX14TC + df$OHX15TC + df$OHX16TC +  df$OHX17TC + df$OHX18TC + df$OHX19TC + df$OHX20TC +  df$OHX21TC + df$OHX22TC + df$OHX23TC + df$OHX24TC +  df$OHX25TC + df$OHX26TC + df$OHX27TC + df$OHX28TC + df$OHX29TC + df$OHX30TC + df$OHX31TC + df$OHX32TC)

table(df$num_teeth)
summary(df$num_teeth)
sd(df$num_teeth)
IQR(df$num_teeth)

#creating number of caries variable
df$num_caries <- (df$OHX02CTC + df$OHX03CTC + df$OHX04CTC + df$OHX05CTC + df$OHX06CTC + df$OHX07CTC + df$OHX08CTC + df$OHX09CTC + df$OHX10CTC + df$OHX11CTC + df$OHX12CTC + df$OHX13CTC + df$OHX14CTC + df$OHX15CTC + df$OHX18CTC + df$OHX19CTC + df$OHX20CTC + df$OHX21CTC + df$OHX22CTC + df$OHX23CTC + df$OHX24CTC + df$OHX25CTC + df$OHX26CTC + df$OHX27CTC + df$OHX28CTC + df$OHX29CTC + df$OHX30CTC + df$OHX31CTC)

table(df$num_caries)
summary(df$num_caries)
sd(df$num_caries)
IQR(df$num_caries)

table(df$SMQ020)
table(df$DBQ700)
table(df$DCOMBO)
table(df$HSD010)

```

# Data Exploration and Analysis

## Best Subset Selection and Validation Set Approach

```{r}
set.seed(1)
# split data into test and training sets
train <- sample(nrow(df), nrow(df)/2)
test=(-train)

regfit.best=regsubsets(DPQ_score ~ num_caries+num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, df[train,])
test.mat=model.matrix(DPQ_score~num_caries+num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, df[test,])
val.errors=rep(1:6)
for (i in 1:6) {
  coefi=coef(regfit.best, id=i)
  pred=test.mat[,names(coefi)]%*%coefi
  val.errors[i]=mean((df$DPQ_score[test]-pred)^2)
}
val.errors
which.min(val.errors)
plot(val.errors,xlab = "Number of Variables", ylab = "Test MSE", main = "Graph 1", pch = 19, type = "b")
points(5,val.errors[5], col="red",cex=2,pch=20)
coef(regfit.best, 5)

# now on the full dataset 
regfit.best=regsubsets(DPQ_score ~ num_caries+num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, df)
coef(regfit.best, 5)
```

## Decision Tree

```{r}
set.seed(1)
train <- sample(nrow(df), nrow(df)/2)
test <- -train
df.train <- df[train,]
df.test <- df[test,]

tree.df <- tree(DPQ_score ~ num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, data=df.train)
summary(tree.df)
plot(tree.df)
text(tree.df, pretty = 0)
```

## Linear, Lasso, and Ridge Regression

```{r}
# Linear
set.seed(1)
train <- sample(nrow(df), nrow(df)/2)
lm.fit = lm(DPQ_score ~ num_teeth + SMQ020 + DBQ700 + HSD010 + DCOMBO, df, subset=train)
summary(lm.fit)
attach(df)
mean((DPQ_score-predict(lm.fit, df))[-train]^2)

#Ridge
x=model.matrix(DPQ_score ~ num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, df)[,-1]
y=df$DPQ_score

grid=10^seq(10,-2,length=100)
ridge.mod=glmnet(x,y,alpha=0,lambda=grid)
predict(ridge.mod,s=50,type="coefficients")[1:5,]
set.seed (1)
train=sample(1:nrow(x), nrow(x)/2)
test=(-train)
y.test=y[test]
ridge.mod=glmnet(x[train,],y[train],alpha=0,lambda=grid, thresh =1e-12)
ridge.pred=predict(ridge.mod,s=4,newx=x[test,])
mean((ridge.pred-y.test)^2)

#Lasso
lasso.mod=glmnet(x[train ,],y[train],alpha=1,lambda=grid)
set.seed (1)
cv.out=cv.glmnet(x[train ,],y[train],alpha=1)
plot(cv.out)
bestlam=cv.out$lambda.min
bestlam
#lambda = 0.007
lasso.pred=predict(lasso.mod,s=bestlam ,newx=x[test,])
mean((lasso.pred-y.test)^2)
```

## Generating Binary DEP Variable

```{r}
df$DEP = df$DPQ_score
df$DEP <- cut(df$DPQ_score,breaks=c(-Inf,8, Inf), labels = c("Non-Dep", "Dep"))
summary(df$DEP)
table(df$DEP)
```

## Logistic Regression, LDA, QDA, KNN

```{r}
#logistic regression
glm.fit = glm(DEP ~ num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, data= df, family='binomial')
summary(glm.fit)
glm.prob = predict(glm.fit, type = 'response')
glm.predict1 = ifelse(glm.prob > 0.5, 'Dep', 'Non-Dep')
table(glm.predict1, df$DEP)
mean(glm.predict1 == df$DEP)

#ROC curve
thres_seq <- seq(from = 0.14, to = 0.36, by = 0.01)
len <- length(thres_seq)
FPR_seq <- rep(0,len)
TPR_seq <- rep(0,len)
ypred <- predict(glm.fit, df, type="response")
for(j in 1:len){
  yhat <- ypred > thres_seq[j]
  mat <- table(df$DEP, yhat)
  FPR_seq[j] <- mat[1,2]/(mat[1,1]+mat[1,2])
  TPR_seq[j] <- mat[2,2]/(mat[2,1]+mat[2,2])
}
plot(FPR_seq, TPR_seq, xlab ='FPR',ylab = 'TPR', main = "ROC Curve")

# LDA
lda.fit.dep=lda(DEP~num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, data=df);
lda.fit.dep
lda.predict.dep = predict(lda.fit.dep, df)
lda.class = lda.predict.dep$class;
table(lda.class, df$DEP)
mean(lda.class != df$DEP)
mean(lda.class == df$DEP)
# 0.1704331
# 0.8295669

#QDA
qda.fit.dep=qda(DEP~num_teeth+SMQ020+DBQ700+HSD010+DCOMBO, data=df);
qda.fit.dep
qda.predict.dep = predict(qda.fit.dep, df)
qda.class = qda.predict.dep$class;
table(qda.class, df$DEP)
mean(qda.class != df$DEP)
mean(qda.class == df$DEP)
# [1] 0.1935028
# [1] 0.8064972

#knn
num_caries <- list(df$num_caries) 
num_teeth <- list(df$num_teeth)
DBQ700 <- list(df$DBQ700)
DCOMBO <- list(df$DCOMBO)
SMQ020 <- list(df$SMQ020)
HSD010 <- list(df$HSD010)
set.seed(1)
train = (df$SEQN < 100592); 

train.X = cbind(df$num_caries, df$num_teeth, df$DBQ700, df$DCOMBO, df$HSD010)[train, ]
test.X = cbind(df$num_caries, df$num_teeth, df$DBQ700, df$DCOMBO, df$HSD010)[!train, ]
train.dep = df$DEP[train]
dep.nt = df$DEP[!train]
knn.train.pred = knn(train.X,train.X, train.dep, k=1)
mean(knn.train.pred != train.dep) ##Training Error
mean(knn.train.pred == train.dep)

knn.test.pred = knn(train.X,test.X, train.dep, k=1)
mean(knn.test.pred != dep.nt) ##Test Error
mean(knn.test.pred == dep.nt)

```



